% This file was converted to LaTeX by Writer2LaTeX ver. 1.0.2
% see http://writer2latex.sourceforge.net for more info
\documentclass[letterpaper]{article}
\usepackage[ascii]{inputenc}
\usepackage[T3,T1]{fontenc}
\usepackage[english]{babel}
\usepackage[noenc]{tipa}
\usepackage{tipx}
\usepackage{amsmath}
\usepackage{amssymb,amsfonts,textcomp}
\usepackage[top=0.9839in,bottom=0.7874in,left=0.9839in,right=0.9839in,nohead,nofoot]{geometry}
\usepackage{array}
\usepackage{hhline}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, filecolor=blue, urlcolor=blue}
% Text styles
\newcommand\textstyleapplestylespan[1]{#1}
% Footnote rule
\setlength{\skip\footins}{0.0469in}
\renewcommand\footnoterule{\vspace*{-0.0071in}\setlength\leftskip{0pt}\setlength\rightskip{0pt plus 1fil}\noindent\rule{0.25\columnwidth}{0.0071in}\vspace*{0.0398in}}
% Pages styles
\makeatletter
\newcommand\ps@Standard{
  \renewcommand\@oddhead{}
  \renewcommand\@evenhead{}
  \renewcommand\@oddfoot{}
  \renewcommand\@evenfoot{}
  \renewcommand\thepage{\arabic{page}}
}
\makeatother
\pagestyle{Standard}
\title{}
\author{Ulrike}
\date{2012-05-16}
\begin{document}
\clearpage\setcounter{page}{1}\pagestyle{Standard}
{\upshape
\textbf{Bibliography}}

{\upshape
\textsc{Atkins, Sue, Jeremy Clear \& Nicholas Ostler} (1992). {\textquotedblleft}Corpus design criteria{\textquotedblright}. In \textit{Literary and Linguistic Computing. }Vol. 7, Nr. 1\textit{ }}

{\upshape
\textsc{Clancy, Brian} (2010). {\textquotedblleft}Building a corpus to represent a variety of language{\textquotedblright}. In Anne O{\textquotesingle}Keeffe and Michael McCarthy. \textit{The Routlege Handbook of Corpus Linguistics. }Abingdon: Routledge, pp. 80-92.}

{\upshape
\textsc{Friedl, Jeffrey E. F. (2006). }\textit{Mastering regular expressions. }Beijing etc.: O{\textquoteleft}Reilly.}

{\upshape
\textsc{Hengeveld, Kees }and\textsc{ Jan Rijkoff (2005). {\textquotedblleft}}Mundari as a flexible language{\textquotedblright}. \textit{Linguistic Typology. }Vol. 9-3, pp. 406-431.}

{\upshape
\textsc{Hengeveld, Kees, Jan Rijkoff \& Anna Siewierska} (2004). {\textquotedblleft}Part-of-speech systems and word order{\textquotedblright}. \textit{Journal of Linguistics} 40, pp. 527-570.}

{\upshape
\textsc{McEnery, Tony \& Andrew Hardie } (2012). \textit{Corpus Linguistics. }Cambridge: CUP.}

{\upshape
\textsc{Mosel, Ulrike} (1977). \textit{Tolai Texts}, Kivung. Journal of the\textit{ }Linguistic Society of Papua New Guinea. Port Moresby. Vol.. 10, pp. XIV + 175. }

{\upshape
\textsc{Mosel, Ulrike} (1980). \textit{Tolai and Tok Pisin}. The influence of the substratum on the development of New Guinea Pidgin. Pacific Linguistics B - 73. Canberra: : Australian National University Press.}

{\upshape
\textsc{Mosel, Ulrike} (1984). \textit{Tolai syntax and its historical development}. Pacific Linguistics B - 73. Canberra A.N.U. Press.}

{\upshape
\textsc{Mosel, Ulrike}. 2006a. {\textquoteleft}The art and craft of writing grammars{\textquoteright}, in Felix Ameka, Alan Dench and Nicholas Evans. \textit{Catching language. }The standing challenge of grammar writing, 41-68.\textit{ }}

{\upshape
\textsc{Mosel, Ulrike} 2006b. Fieldwork and community language work, in Jost Gippert, Nikolaus Himmelmann and Ulrike Mosel (eds.) \textit{Essentials of language documentation. }Berlin, New York: Mouton de Gruyter, 67-85. }

{\upshape
\textsc{Mosel, Ulrike} (forthcoming). Corpus linguistic and documentary approaches in writing a grammar of a previously undescribed language. In: Toshihide Nakayama \& Karen Rice. \textstyleapplestylespan{\textit{Practical and Methodological Issues in Grammar Writing.}}}

{\upshape
\textsc{Mosel, Ulrike \& Even Hovdhaugen} (1992). \textit{Samoan Reference Grammar}. Oslo: Scandinavian University Press.}

{\upshape
\textsc{Mosel, Ulrike, Enoch Horai Magum, Shalom Magum, Joyce Maion, Naphtaly Maion, Jessika Reinig, Ruth Siimaa Rigamu, Ruth Saovana Spriggs, \& Yvonne Thiesen} (2007). \textit{The Teop Language Corpus.} http://www.mpi.nl/dobes/projects/teop }

{\upshape
\textsc{Mosel, Ulrike} with \textsc{Yvonne Thiesen} (2007) \textit{Teop sketch grammar.} \href{http://www.mpi.nl/DOBES/projects/teop}{www.mpi.nl/DOBES/projects/teop} and \hyperlink{download}{http://www.linguistik.uni-kiel.de/mosel\_publikationen.htm\#download} }

{\upshape
\textsc{Nordhoff, Sebastian }(2008). {\textquotedblleft}Electronic reference grammars for typology: Challenges and solutions{\textquotedblright} \textit{Language Documentation and Conversation, }2:296-324.}

{\upshape
\textsc{Sankoff, Gillian (1993). {\textquotedblleft}}Focus in Tok Pisin. In Francis Byrne \& Donald Winford (eds.) \textit{Focus and grammatical relations in Creole languages.} Amsterdam/Philadelphia: John Bejamins Publishing Company, pp. 117-140.}

{\upshape
\textsc{Schwartz, Maricia L., Ruth Saovana Spriggs, Ulrike Mosel, Ruth Siimaa Rigamu, Naphtaly Maion, \& Jeremiah Vaabero }(2007). The Teop Lexical Database. \textsc{ } http://www.linguistik.uni-kiel.de/Teop\_Lexical\_Database\_May07.pdf}

{\upshape
\textsc{Stefanowitsch, Anatol \& Stefan Gries} (2009). Corpora and grammar. In Anke L\"udeling \& Meja Kyt\"o (eds.). \textit{Corpus linguistics. An international Handbook. }HSK Vol. 29.2. Berlin \& New York: Walter de Gruyter, pp. 933-952.}

{\upshape
\textsc{Whalen, Doug H. (2004). }How the study of endangered languages will revolutionize linguistics. In Piet van Sterkenburg (ed.). \textit{Linguistics today - facing a greater challenge. }Amsterdam/Philadelphia: John Benjamins Publishing Company, pp.321-342.}
\end{document}
