\chapter{Pragmatics}\label{sec:Pragmatics}
After we have seen which are the elements and constructions of SLM grammar, and how different meanings can be expressed by them, we now turn to the different communicative goals that speakers may want to achieve through the use of these units. We will first discuss different types of speech acts (requesting confirmation, requesting action, answering, etc.) \funcref{sec:pragm:Speechacts} and then turn to the ways how speech acts can be nuanced with regard to force and politeness \funcref{sec:pragm:Blendinginthesocialtissue}.

\section{Speech acts}\label{sec:pragm:Speechacts}


\subsection{Providing information}\label{sec:pragm:Providinginformation}
If a speaker wants to provide information, he uses an assertion.
Assertions have been dealt with extensively in the preceding chapters, and all relevant information can be taken from there.

Speakers do not only want to provide information. Sometimes, they also want the hearers to do something, e.g. provide missing information or perform a certain action \citep[319]{Givon2001b}. They might also want to express their agreement with the hearer or someone else to perform a certain action. This will be dealt with in the following sections.

\subsection{Requesting information}\label{sec:pragm:Requestinginformation}
Requesting information can be divided into two subtypes: requesting information about the truth of a certain proposition, and requesting information about the content of a certain proposition. The former type is called `polar question' and the latter, `content question'.

Content questions are formed in SLM by means of an interrogative clause for content \formref{sec:cls:InterrogativeclauseWH}, which resembles a normal clause, but replaces the NP or PP queried for by an interrogative pronoun like \trs{saapa}{who}.  

A special  type of content question is concerned with  metalinguistic information about a translation of a word into another language. For this, the dative marker \em =nang \em is used. The interrogative pronoun is either \em aapa \em as in \xref{ex:pragm:reqinf:metaling:aapa} or \em aapayang \em as in \xref{ex:pragm:reqinf:metaling:aapayang}.

\xbox{16}{
\ea\label{ex:pragm:reqinf:metaling:aapa}
\gll {\em ``Simple''}=\textbf{nang} \textbf{aapa} arà-biilang. \\ % bf
     simple=\textsc{dat} what \textsc{non.past}-say  \\
    `What do you say for "simple"?' (B060115cvs01)
\z
}


\xbox{16}{
\ea\label{ex:pragm:reqinf:metaling:aapayang}
\gll {\em ``Terrorist''}=\textbf{nang}   \textbf{aapayang} arà-biilang,    mlaayu=dering. \\ % bf
      terrorist=\textsc{dat} what \textsc{non.past}-say Malay=\textsc{abl}\\
    `How do you say "terrorist" in Malay?' (K051206nar02)
\z
}

The second possibility are polar questions, which are formed by a YN-interrogative clause \formref{sec:cls:Interrogativeclauseclitic}, or by a declarative clause followed by the tag \trs{bukang}{isn't it?}

The clitic \em =si \em is neutral as to the expected answer. In example \xref{ex:pragm:reqinf:neutral}, the speaker is not sure whether she has the obligation to tell her age or not. If a positive answer is expected, the declarative sentence followed by  \trs{bukang}{isn't it} is used \xref{ex:pragm:reqinf:bukang1}\xref{ex:pragm:reqinf:bukang2}.


\xbox{16}{
\ea\label{ex:pragm:reqinf:neutral}
\gll Se=pe uumur masà-biilan=si? \\ % bf
 1=\textsc{poss} age must-tell=\textsc{interr}\\
`Do I have to tell my age?' (B060115prs01)
\z
}

\xbox{16}{
\ea\label{ex:pragm:reqinf:bukang1}
\gll Itthu    saala, bukang?\\ % bf
\textsc{dist} wrong \textsc{neg.nonv}\\
`That is wrong, isn't it?' (K060116nar11)
\z
}

\xbox{16}{
\ea\label{ex:pragm:reqinf:bukang2}
\gll Kandi=ka ithu Thuan Skiilan aada, bukang? \\ % bf
      Kandy=\textsc{loc} \textsc{dist} Thuan Skiilan exist \textsc{tag}\\
    `Now there is this Thuan Skiilan in Kandy, you know?' (B060115cvs04)
\z
}


When expecting a negative answer, a negative declarative sentence is used, followed by \em =si\em.

\xbox{16}{
\ea\label{ex:pragm:reqinf:neg=si}
\gll Puaasa muusing thàrà-duuduk=si? \\ % bf
fasting season \textsc{neg}-stay=\textsc{interr}\\
`You were not here during the fasting season, were you?' (B060115cvs03)
\z
} 

Questions are also marked by a particular intonation contour (Section \ref{sec:phon:Questioncontour}, p. \pageref{sec:phon:Questioncontour}). If that contour is clear, the segmental marking (\em =si\em) can be left out from polar questions.
 
\subsection{Requesting confirmation}\label{sec:pragm:Requestingconfirmation}
Requesting confirmation of the expected truth value has been discussed above in the preceding section. Another kind of confirmation which can be requested is whether the hearer is actually following the words of the speaker. This can be done by \trs{thaau=si}{you know?}. The hearer can then confirm verbally or non-verbally, and the speaker can continue his turn.

\xbox{16}{
\ea\label{ex:pragm:reqconf:thausi}
\gll Bannyak pukurjan anà-kirja    soojor pada-saama,      \textbf{thaau=si}  soojor? \\
     much work \textsc{past}-do European \textsc{pl}=\textsc{comit} know=\textsc{interr} European  \\
    `He worked together with ``soojors''. You know ``soojor''?' (K061026prs01)
\z
}

Another possibility to request confirmation is \em =jona \em \formref{sec:morph:=jona}.


\xbox{16}{
\ea\label{ex:pragm:reqconf:jona}
\gll Punnu mlaayu pada kàthaama {\em English}=\textbf{jona} anthi-oomong. \\
       many Malay \textsc{pl} earlier English=\textsc{phat} \textsc{irr}=speak \\
    `Many Malays would speak English, y'know,  in former times.'  (K051222nar06)
\z      
}\\ 


\subsection{Requesting permission}\label{sec:pragm:Requestingpermission}
Permission is requested by adding the interrogative clitic \em =si \em \formref{sec:morph:=si} to a verb marked by either the infinitive \em mà- \em \formref{sec:morph:ma-} or the adhortative \em marà- \em \formref{sec:morph:mara-}.


\xbox{16}{
\ea
\gll See \textbf{mà}-maakang=\textbf{si}? \\
     \textsc{1s} \textsc{inf}-eat=\textsc{interr}  \\
    `Shall I eat?' (K081106eli01) 
\z
}


\xbox{16}{
\ea
\gll Kithang \textbf{marà}-maakang=\textbf{si}? \\
      \textsc{1pl} \textsc{adhort}-eat=\textsc{interr}\\
    `Shall we eat?' (K081106eli01) 
\z
}

 
\subsection{Requesting action}\label{sec:pragm:Requestingaction}
Next to providing and requesting information, speakers also sometimes want other people to perform certain actions, or to refrain from performing certain actions.
The most straightforward way to request action is an imperative construction \formref{sec:cls:Imperativeclause}.

\xbox{16}{
 \ea\label{ex:pragm:reqact:pos}
   \gll Binthan {\em auntie}=ka    caanya, Binthan {\em auntie}=yang   konnyong panggel. \\  % bf
    Binthan auntie=\textsc{loc} ask, Binthan auntie=\textsc{acc} few call \\
`Ask auntie Binthan, call Binthan auntie' (K060116nar06)
\z
}
 

Milder than the preceding utterance is the use of \trs{maau}{want} \formref{sec:wc:(ka)mau(van)}, as in the following examples.

\xbox{16}{
 \ea\label{ex:pragm:reqact:maau:mosque}
\gll Baapa=nang {\em mosque}=nang mà-pii \textbf{maau}. \\
 father=\textsc{dat} mosque=\textsc{dat} \textsc{inf}-go want\\
`Father, you should go to the mosque/ Father, you want to go to the mosque.' (B060115nar04)
\z
}

\xbox{16}{
 \ea\label{ex:pragm:reqact:maau:recipe}
\gll Manis-an=nang mà-thaaro guula \textbf{maau} gula paasir konnyong \textbf{maau}. \\
 sweet-\textsc{nmlzr}=\textsc{dat} \textsc{inf}-put sugar want sugar sand few want\\
`In order to make it sweet, you need sugar, you need a bit of crystal sugar.' (B060115rcp02)
\z
}

The negation of \em maau\em, \em thussa \em \formref{sec:wc:thussa}, can be used for negative commands, as in \xref{ex:cl:imp:neg:thussa}.

\xbox{16}{
\ea\label{ex:cl:imp:neg:thussa}
\gll `Thussa mà-thaakuth', Buruan su-biilang. \\ % bf
      \textsc{neg.imp} \textsc{inf}-fear bear \textsc{past}-say\\
    ` ``Don't be afraid,''  said Bear.'/` ``I don't want you to be afraid'', said Bear.'  (K070000wrt04)
\z      
}\\ 

Another more subtle way is to use \trs{lorangnang ... (thàr)boole}{you can(not) ...}  \formref{sec:wc:therboole} \xref{ex:pragm:reqact:therboole}, or the verbal prefix \trs{masthi}{must} \formref{sec:morph:masthi-} with an unexpressed referent \xref{ex:pragm:reqact:masthi}. Both imply that it would be preferable that the action be (not) performed, but lack coercive power (in the case of \em (thàr)boole\em) or the direct expression of the addressee (in the case of \em masthi\em).

\xbox{16}{
 \ea\label{ex:pragm:reqact:therboole}
\gll Ithu-kapang umma-baapa  su-biilang     lorang=nang   kaaving \textbf{thàrboole}. \\ % bf
     \textsc{dist}=when mother-father \textsc{past}-say \textsc{2pl}=\textsc{dat} marry cannot  \\
    `Then the parents said: ``You cannot marry/Don't marry.''\,' (K051220nar01)
\z
}

\xbox{16}{
 \ea\label{ex:pragm:reqact:masthi}
\gll Minnyak klaapa ini raa\u mbuth=dering \textbf{masà}-goosok. \\ % bf
      coconut.oil coconut \textsc{prox} hair=\textsc{abl} must-rub \\
`You must rub coconut oil (over the itching) with (human) hair.' (K060103cvs02)
\z
}

Adhortatives can be formed with \em marà- \em \formref{sec:morph:mara-}.

\xbox{16}{
 \ea\label{ex:pragm:reqact:mara}
\gll Kitham \textbf{marà}-maayeng. \\
     \textsc{1pl} \textsc{adhort}-play  \\
    `Let's play.' (K081104eli06)
\z
}




\subsection{Answering}\label{sec:pragm:Answering}
Answering affirmatively is done by \em iiya \em \xref{ex:pragm:answer:yes} or \em saya\em. Negative answers are given by \em thraa\em, regardless of predicate type \xref{ex:pragm:answer:no}.


\xbox{16}{
 \ea\label{ex:pragm:answer:yes}
\ea
\gll SN: Butthul {\em sportsman}, bukang? \\ % bf
     { } correct sportsman, \textsc{tag}\\
    `They are good sportsmen, aren't they?' 
\ex
\gll SLM: \textbf{iiya} \textbf{iiya}, {\em sportsman}.    \\
     { } Yes, yes, sportsman  \\
    `Yes indeed, they are good sportsmen.'  \\
\el
\ex
\gll SN: {\em Football} aada? \\ % bf
     {  } Football exist  \\
    `Is there football being played in Badulla.'  
\ex
\gll SLM: \textbf{Iiya} {\em football} aada. \\ % bf
     {  }  yes football exist \\
    `Yes, there is football.' (B060115cvs01)
\z
\z
}
 

% 
% \xbox{16}{
% \ea\label{ex:pragm:answer:no}
% \gll Ithu=kapang       derang nya-biilang:    \textbf{thraa}, kithang giithu   thama-pii. \\
%       \textsc{dist}-when \textsc{3pl} \textsc{past}-say no \textsc{1pl} like.that \textsc{neg.nonpast}-go \\
%     `Then they said: ``no, we cannot go like that.' '' (K051213nar06)
% \z
% }


It is also possible to repeat the  predicate of the question for affirmative purposes, or to repeat a verbal predicate and add a negative prefix (\em thàrà-/thama-\em) as a negative answer. This can be seen for example in \xref{ex:pragm:answer:no}, where the predicate \trs{kaaving}{marry} is repeated with the negative prefix \em thama- \em in the answer (Why the non-past prefix is chosen here is unclear, though).

\xbox{16}{
 \ea\label{ex:pragm:answer:no}
\ea
\gll Q: Sebastian su-\textbf{kaaving}=si? \\ % bf
     { } Sebastian \textsc{past}-marry=\textsc{interr}?  \\
    `Is Sebastian you married?'
\ex
\gll A: \textbf{Thraa} \textbf{thraa},  Sebastian  thama-\textbf{kaaving}. \\ % bf
      { }  no no Sebastian \textsc{neg.nonpast}-marry \\
    `No, no, Sebastian is not married.'\footnotemark (B060115cvs03)
\z
\z
}
\footnotetext{A second native speaker answers instead of the researcher.}

If the answer to a question reasserts a known referent, this referent is often reduplicated. This is the case in \xref{ex:pragm:answer:repeat}, where the string \em dìkkath \em is present in the question and hence active in discourse. In the answer, this string is present twice, indicating that the speculation of the hearer as to the vicinity of the school has been understood and is indeed correct.

\xbox{16}{
\ea\label{ex:pragm:answer:repeat}
\ea
\gll Q: Skuul dìkkath=si? \\ % bf
 { } school vicinity=\textsc{interr}\\
`Is the school close (expecting yes).' 
\ex
\gll A: Dìkkath dìkkath. \\ % bf
{ } vicinity vicinity\\
`Oh yeah, it is close indeed.' (K081104eli06)
\z
\z
}

\subsection{Allowing}\label{sec:pragm:Allowing}
Allowing is normally done with the modal \trs{boole}{can} \formref{sec:wc:boole}.

\xbox{16}{\ea
\ea
\gll Dadaa, se=dang sinema mà-liiyath=nang bolle=pii=si? \\
     Daddy \textsc{1s=dat} cinema \textsc{inf}-see=\textsc{dat} can=go-\textsc{interr}  \\
    `Daddy, can I go to the cinema?' 
\ex
\gll Iiya, Izi=nang bolle=pii. \\
      yes, Izi=\textsc{dat} can=go \\
    `Yes, you can go, Izi.' (K081104eli06)
\z
\z
}

The lexical verb for `to allow' is \em luppas\em, which also means  `leave (behind)'.

\xbox{16}{
\ea
\gll See Izi=yang arà-luppas sinema mà-liyath=nang mà-pii. \\
     \textsc{1s} Izi=yang \textsc{non.past}-leave cinema \textsc{inf}-watch=\textsc{dat} \textsc{inf}-go  \\
    `I allow you/Izi to go to the cinema.' (K081104eli06)
\z
}

\subsection{Wishing}\label{sec:pragm:...makeawish}
Wishes can be made by invoking Allah. Interestingly, the complement of wishes is marked with the pseudo-prefix \trs{masà-}{must}, instead of an irrealis marker or an infinitive, which would have been expected.


\xbox{16}{
\ea
\gll Yaa, Allah, se arà-mintha se=dang itthu pukuran masà-daapath katha. \\
      Yeah, Allah, \textsc{1s} \textsc{non.past}-wish \textsc{1s=dat} \textsc{dist} job must-get \textsc{quot}\\
    `Allah, I wish I could get that job.' (K081104eli06)
\z
}

% 
% \xbox{16}{
% \ea
% \gll Se aràmintha lorang se=dang masà-daapath katha. \\
%        \\
%     `.' (K081104eli06)
% \z
% }
% 
% 
% \xbox{16}{
% \ea
% \gll Se arà mintha lorang se=ppe masà/?anthi jaadi katha. \\
%        \\
%     `.' (K081104eli06)
% \z
% }


Wishes can also be indicated by \trs{incalla}{if God wills}.


\xbox{16}{
\ea\label{ex:pragm:wish:incalla}
\gll Incalla   [lai     thaau sudaara sudaari pada]=ka    bole=caanya    ambel [nya-gijja    lai     saapa=kee  aada]=si    katha. \\
      Hopefully other know brother sister \textsc{pl}=\textsc{loc} can-ask take \textsc{past}-make other who=\textsc{simil} exist=\textsc{interr} \textsc{quot}\\
    `Hopefully, you can enquire from another person you know whether there is someone else who did something.' (N061031nar01)
\z
}


% \subsection{... tell stories}\label{sec:pragm:...tellstories}
% 
% 
% \xbox{16}{
% \ea\label{ex:form:unreferenced}
% \gll Karang liyath, kithang=pe tsunami (pe) atthas[=ka] ini aapa arà(anà)-jaadi katha. \\
%       now see.imp \textsc{1pl}=\textsc{poss} tsunami about=\textsc{loc} \textsc{prox} what \textsc{non.past}-become \textsc{quot}
%     `Now see (hear) what happened with this Tsunami.' (K061127nar03,K081104eli06)% modified
% \z
% }

\subsection{Mirativity}
Mirativity can be expressed by the insertion of \trs{thaausi?}{you know what!?}.

\xbox{16}{
\ea
\gll 
Suda inni moonyeth pada ]\textbf{aapa} \textbf{thaau=si}[ anà-gijja  !\\
 so \textsc{prox} monkey \textsc{pl} what know=\textsc{interr} \textsc{past}-do      \\
    `Do you know what these monkeys then did!?'  (K070000wrt01)
\z      
}\\

Another possibility involves the use of the irrealis marker \em anthi-\em. shown in \xref{ex:disc:mirativity:anthi}.


\xbox{16}{
\ea \label{ex:disc:mirativity:anthi}
\gll Itthu saapa \textbf{anthi-aada}! \\
       \textsc{dist} who \textsc{irr}-exist\\
    `Who could/might that be!'  (K070000wrt04)
\z      
} 


\subsection{Promises}\label{sec:pragm:...promise}
Promises are made with the irrealis marker \em anthi- \em  \formref{sec:morph:anthi-} \xref{ex:pragm:promise:anthi} or with the non-past marker \em arà- \em \formref{sec:morph:ara-} \xref{ex:pragm:promise:sumpa}.

\xbox{16}{
\ea\label{ex:pragm:promise:anthi}
\gll Kithang lorang=nang baaye mliiga \textbf{athi}-kaasi,  mà-kaaving panthas pompang pada \textbf{athi}-kaasi,  duvith \textbf{athi}-kaasi. \\
      \textsc{1pl} \textsc{2pl}=\textsc{dat} good palace \textsc{irr}-give \textsc{inf}-marry beautiful girl \textsc{pl} \textsc{irr}-give money \textsc{irr}-give \\
    `We will give you beautiful palaces and women to marry and money.' (K051213nar06)
\z
}

\xbox{16}{
\ea\label{ex:pragm:promise:sumpa}
\gll See arà-\textbf{sumpa}  paanas muusing dhaathang=thingka see siini=dering \textbf{arà}-pii. \\
     \textsc{1s} \textsc{non.past}-promise hot time come=middle \textsc{1s} here=\textsc{abl} \textsc{non.past}-go  \\
    `I promise, I will leave when summer will have come.' (K070000wrt04)
\z
}

\subsection{Explanations}\label{sec:pragm:...explain}
Explanations are often structured along the pattern of statement, rhetorical question, repetition of statement. The following stretch of discourse by one speaker is in English, but the pattern is also found in SLM even though it is not in the corpus.

 
\xbox{16}{
\ea
\ea  Two miles we have to walk.   
\ex How many miles? 
\ex
\gll {\em  Two} {\em mile} duuva kaayu kithang masà-pii. \\
     two mile two mile \textsc{1pl} must-go  \\
    `Two miles we had to walk.' (K051213nar03)
\z
\z
}

\subsection{Greeting people}\label{sec:pragm:...greetpeople}
While the other native ethnic groups in Sri Lanka do normally not express greetings verbally, this is different for the Malays.\footnote{Tamil \em vanakkam \em and Sinhala \em Aayu boovan \em are only used for special occasions. One Sinhala informant reported that he could not remember when he had last used the greeting. Things are a bit different for Moors, who use \em Salam aleikum \em on a regular basis.}
The most common greeting is \em slaamath\em, which can be uttered at arrival or departure, and also on the telephone. It can be complemented by the period of the day \trs{slaamath paagi}{good morning}, \trs{slaamath soore}{good afternoon/evening}, and \trs{slaamath maalang}{good night}{}. More informal greetings do not exist, but English \em hello \em can be used for such purposes.

More formal is \trs{slaamath dhaathang}{welcome}, which can be uttered at the same occasions as its English counterpart.


\subsection{Taking leave}\label{sec:pragm:...takeleave}
When leaving, one can use \em slaamath \em as described above, or one can use \trs{spi dhaathang}{go and come}. This is modelled on an analogous construction in Sinhala and Tamil. The rationale for this is that it is considered unfortunate to say \em I leave \em because that might mean departure from this world. Therefore, one has to specify that one will return, and this is done by adding the \trs{dhaathang}{come} part. Optionally, the \em spi-\em part can be left out, which leaves the curious situation that you can say \em I am coming \em to indicate you are leaving, as in the following example.

\xbox{16}{
\ea\label{ex:pragm:unreferenced}
\gll Se=dang arà-{\em late}, bukang, \textbf{see} \textbf{arà-dhaathang}. \\
      \textsc{1s=dat} \textsc{non.past}-late \textsc{tag} \textsc{1s} \textsc{non.past}-come \\
    `I am getting late, aren't I, goodbye.' (B060115cvs08)
\z
}

\subsection{Thanking}\label{sec:pragm:...thank}
Just as with greetings, Sinhalese and Tamils do not use verbal means for this function, but Malays do. The normal way of thanking is \trs{thriima kaasi}{thank give}.

 
\xbox{16}{
\ea
\gll (Bannyak) thriima kaasi. \\
      much thank give \\
    `Thank you very much.' (ubiquitous)
\z
}

% \subsection{... insist}\label{sec:pragm:...insist}
% \subsection{... get angry}\label{sec:pragm:...getangry}
% 
% 
% \xbox{16}{
% \ea
% \gll. \\
%        \\
%     `.' (nosource)
% \z
% }



\subsection{Asking a favour}\label{sec:pragm:...askafavour}
Asking a favour is normally done with an interrogative construction involving \em =si\em. Given that the favour is typically beneficial to the asker, the vector verb \trs{kaasi}{give} is often used then.

\xbox{16}{
\ea\label{ex:pragm:askfavour:ambel}
\gll Mamaa, se=dang itthu=yang anthi-ambe \textbf{kaasi=si}? \\
      mum \textsc{1s=dat} \textsc{dist}=\textsc{acc} \textsc{irr}-take give=\textsc{interr}\\
    `Mummy, can you buy that for me?' (K081104eli06)
\z
}


\xbox{16}{
\ea
\gll Mamaa, se=dang nyaari daaging athi-maasak \textbf{kaasi=si}?   \\
     mum, \textsc{1s=dat} today meat \textsc{irr}-cooked give=\textsc{interr}  \\
    `Mama, can you cook beef for me today?' (K081104eli06)
\z
}

The asking of favours involves polite pronouns. It is even possible to use double marked plural forms like \em loram pada \em with singular reference then.\footnote{Tamil also uses plural pronouns for polite reference \citep[100]{Lehmann1989}.}

\xbox{16}{
\ea\label{ex:pragm:askfavour:GFG}
\gll Saayang, \textbf{loram} \textbf{pada}=sesama nyaari {\em Galle} {\em Face} mà-pii arà-dhaathang=si? \\
     love \textsc{2pl} \textsc{pl}=\textsc{comit} today Galle Face \textsc{inf}-go \textsc{non.past}-come=\textsc{interr}   \\
    `Darling, shall I come and take you to the Galle Face Green [a favourite hangout for young couples in Colombo]?' (K081104eli06)
\z
}



% \xbox{16}{
% \ea
% \gll {\em Auntie}, se arà mintha incayang se=dang masà daapath katha. \\
%        \\
%     `.' (K081104eli06)
% \z
% }

\xref{ex:pragm:askfavour:auntie} gives a request about a third person.

\xbox{16}{
\ea\label{ex:pragm:askfavour:auntie}
\gll {\em Auntie}, se arà-mintha incayang se=ppe masà-jaadi katha. \\
    auntie \textsc{1s} \textsc{non.past}-ask 3s.polite \textsc{1s=poss} must-become \textsc{quot}   \\
    `Madam, may I ask for the hand of your daughter.' (K081104eli06)
\z
}



\subsection{Granting a favour}\label{sec:pragm:...grantafavour}
Granting a favour is done by \trs{iiya}{yes} or \trs{butthul}{correct, OK}. \xref{ex:pragm:grantfavour:ambel} is the positive answer for request \xref{ex:pragm:askfavour:ambel}.

\xbox{16}{
\ea\label{ex:pragm:grantfavour:ambel}
\gll Iiya, butthul, se athi-ambel kaasi. \\
     yes correct \textsc{1s} \textsc{irr}-take give  \\
    `Yes, I will buy it for you.' (K081104eli06) %butthul sociolgstc
\z
}

The positive reply to \xref{ex:pragm:askfavour:GFG} is \xref{ex:pragm:grantfavour:GFG}. Note the use of the terms \trs{saayang}{love} for the girl in \xref{ex:pragm:askfavour:GFG} and \trs{jiiva}{life} for the boy in  \xref{ex:pragm:grantfavour:GFG}.

\xbox{16}{
\ea\label{ex:pragm:grantfavour:GFG}
\gll Iiya, jiiva, se  thi-dhaathang. \\
      yes life \textsc{1s} \textsc{irr}-come \\
    `Yes, dear, I will come.' (K081104eli06)
\z
}



\subsection{Declining a favour}\label{sec:pragm:...declineafavour}
The negative answer to \xref{ex:pragm:askfavour:ambel} is \xref{ex:pragm:declinefavour:ambel}, the negative answer to \xref{ex:pragm:askfavour:GFG} is \xref{ex:pragm:declinefavour:GFG}, the negative answer to \xref{ex:pragm:askfavour:auntie} is \xref{ex:pragm:declinefavour:auntie}.

\xbox{16}{
\ea\label{ex:pragm:declinefavour:ambel}
\gll Thraa, se loran=nang thama-ambel kaasi. \\
    \textsc{neg} \textsc{1s} \textsc{2pl}=\textsc{dat} \textsc{irr}-take give   \\
    `No, I won't buy it for you.' (K081104eli06)
\z
}


% \xbox{16}{
% \ea
% \gll thraa, se(re)ka duvith thraa. \\
%        \\
%     `.' (K081104eli06)
% \z
% }


\xbox{16}{
\ea\label{ex:pragm:declinefavour:GFG}
\gll Thookal=si? \\
      fool=\textsc{interr}\\
    `Are you crazy?/Get lost!' (K081104eli06)
\z
}

\xbox{16}{
\ea\label{ex:pragm:declinefavour:auntie}
\gll Thookal, itthu kalu, thama-jaadi! \\
     fool \textsc{dist} if \textsc{neg.irr}-become  \\
    `You fool! As for that, that will never happen!' (K081104eli06)
\z
}





\subsection{Telling your age}\label{sec:pragm:...tellyourage}
There are two ways to tell your age, one modelled on the European pattern and one on the Lankan pattern. The European pattern is stating the number of your last birthday, while the Lankan way is stating the number of your next birthday with the verb \trs{gìnnap}{complete}. Both cases put the argument in the dative case, in the examples \em sedang\em.



\xbox{16}{
\ea\label{ex:pragm:speechact:age:european}
\gll Karang se=dang ìnnam-pulu liima (thaaun) uumur. \\ % bf
     now \textsc{1s=dat} six-ty (year) five age  \\
    `Now I'm sixty-five.' (N060113nar03)
\z
}


\xbox{16}{
\ea\label{ex:pragm:speechact:age:lankan}
\gll Sangke=nyaari se=dang  limapulu  thaaun arà-gìnnap. \\
 until=today \textsc{1s=dat} fifty year \textsc{non.past}-complete\\
`I am in the course of my fiftieth year on earth (= I am 49).' (K060108nar01)
\z
}
 





\subsection{Introducing yourself}\label{sec:pragm:...introduceyourself}
A formal way to introduce yourself is given in \xref{ex:pragm:speechact:introduction}. 

\xbox{16}{
\ea\label{ex:pragm:speechact:introduction}
\gll Sudaara sudaari se=ppe naama Wahida Jamaldiin. \\ % bf
     brother sister \textsc{1s=poss} name Wahida Jamaldeen  \\
    `Brothers and sisters, my name is Wahida Jamaldeen.' (B060115prs05)
\z
}



\subsection{Expressing  annoyance}
Annoyance is not often expressed openly. The following possibility was gathered through elicitation.

\xbox{16}{
\ea
\gll Cumma duuduk-la! \\
     idle stay-\textsc{imp}  \\
    `Mind your own business!' (K081105eli02)
\z
}


\subsection{Ending a conversation}\label{sec:pragm:...endaconversation}
An explicit way to end a conversation is given in \xref{ex:pragm:end:explicit}. The sentences in \xref{ex:pragm:end:implicit1} and \xref{ex:pragm:end:implicit2} are heard very often, but I am not very sure about their pragmatic implications. They might be a polite hint to stop now, or, on the contrary, they might signal desire to continue.

\xbox{16}{
\ea\label{ex:pragm:end:explicit}
\gll Skaarang marà-birthi-king, se=dang lai pukuran aada. \\
     now \textsc{adhort}-stop-\textsc{caus}, \textsc{1s=dat} more work exist  \\
    `Let's stop now, I still have some work to do.' (K081104eli06)
\z
}


\xbox{16}{
\ea\label{ex:pragm:end:implicit1}
\gll Lai aapa aada?\\
      more what exist \\
    `What else?' (K081104eli06)
\z
}


\xbox{16}{
\ea\label{ex:pragm:end:implicit2}
\gll Lai aapa=ke (mà-oomong) aada=si? \\
     more what=\textsc{simil} \textsc{inf}-talk exist=\textsc{interr}  \\
    `Is there something else to say?' (K081104eli06)
\z
}


\subsection{Hesitation}
Hesitation is marked by the proximal deictic \em inni\em, as in \xref{ex:pragm:hesitation}, when looking for the right word.


\xbox{16}{
\ea\label{ex:pragm:hesitation}
\gll laile itthu=yang      arà-..\textbf{inni}..-iingath-kang   inni     police=dering \\
      still \textsc{dem.dist}=\textsc{acc} \textsc{non.past}-..\textsc{hesit}..-think-\textsc{caus} \textsc{dem.prox} police=\textsc{abl}   \\
`Still the police bring into memory this story.' (K051206nar02)
\z
}


\section{Blending in the social tissue}\label{sec:pragm:Blendinginthesocialtissue}
This section is mainly concerned with politeness. The main expression of politeness is through personal pronous (see Table \ref{tab:PersonalPronouns} on page \pageref{tab:PersonalPronouns}). By choosing the polite form, distance is conveyed, whereas the intimate form conveys closeness. As for bound morphology, the optional imperative suffix \em -la \em  \formref{sec:morph:-la} makes a request more polite, while the impolite imperative suffix \em -de \em \formref{sec:morph:-de} yields a very impolite request , but increases the force of the request.

\subsection{Terms of address}
When speaking to  relatives, the term for the kinship relation may be used instead of the 2$^{nd}$ or 3$^{rd}$ person pronoun \citep[55]{Saldin2001}.  \xref{ex:pragm:address:kin:2:baapa} and \xref{ex:pragm:address:kin:2:aanak} show the use of kin terms as terms of address.


\xbox{16}{
\ea\label{ex:pragm:address:kin:2:baapa}
\gll Hatthu=le jamà-gijja, \textbf{baapa} ruuma=ka duuduk. \\
      \textsc{indef}=\textsc{addit} \textsc{neg.nonfin}-make father house=\textsc{loc} stay \\
    `Father, do not do anything and stay at home!' (B060115nar04)
\z
}


\xbox{16}{
\ea\label{ex:pragm:address:kin:2:aanak}
\ea
\gll Andare aanak=nang su-biilang:\\ % bf
      Andare child=\textsc{dat} \textsc{past}-say \\
    `Andare said to his child:'
\ex
\gll \textbf{Aanak}\\
     child  \\
    ` ``Son, '' '
\ex
\gll 'Lu=ppe        umma su-maathi'     katha  bitharak=apa   asà-naangis   mari. \\ % bf
      \textsc{2s}=\textsc{poss} mother \textsc{past}-die \textsc{quot} scream=after \textsc{cp}-weep come.\textsc{imp}\\
    ` ``come and cry and weep `My mother has died!' '' '
\z
\z
}


The use of kin terms when referring to third person is shown in  \xref{ex:pragm:address:kin:3}, a stretch of discourse about the father of the speaker, who, despite of being introduced early on, is mostly referred to as \em baapa\em. The zeroes indicate instances where the referent is not overtly realized. Also note the reference to \trs{aade}{younger sibling} in a similar sense.
\newpage

\xbox{16}{
\ea\label{ex:pragm:address:kin:3}
\ea
\gll Itthu    blaakang \textbf{baapa}  sajja anà-duuduk. \\
     \textsc{dist} after father only \textsc{past}-saty  \\
    `The father stayed at home alone.'  
\ex
\gll \textbf{Baapa}  asà-duuduk suda giithu samma oorang=pe samma aanak pada=pe  ruuma  \textbf{baapa}  arà-pii. \\
     father \textsc{cp}-stay thus like.that all man=\textsc{poss} all child \textsc{pl}=\textsc{poss} house father \textsc{non.past}-go  \\
    `Father stayed there and then like that father went to all folks', all children's homes.'
\ex
\gll Giithu   asàduuduk     kithang {\em transfer} su-dhaathang     {\em Kandy}=nang. \\ % bf
     like.that cp.stay/from \textsc{1pl} transfer \textsc{past}-come Kandy=\textsc{dat}  \\
    `From there, we were transfered to Kandy.' 
\ex
\gll Kandy=nang   dhaathang=jo    \textbf{baapa}=le       su-dhaathang     {\em Kandy}=nang. \\
     Kandy=\textsc{dat} come=\textsc{emph} father=\textsc{addit} \textsc{past}-come Kandy=\textsc{dat}  \\
    `Only when we came to Kandy did father come to Kandy.' 
\ex
\gll \zero{} kithang=pe     ruuma=ka=le   anà-duuduk, \textbf{aade}=pe  ruuma=ka=le    anà-duuduk. \\
      { } \textsc{1pl}=\textsc{poss} house=\textsc{loc}=\textsc{addit} \textsc{past}-stay younger.sibling=\textsc{poss} house=\textsc{loc}=\textsc{addit} \textsc{past}-stay \\
    `He stayed at our house and he stayed at (my) younger sibling's house.' 
\ex
\gll \zero{} itthu samma         ruuma=ka    asà-duuduk  \\ % bf
     { } \textsc{dist} all house=\textsc{loc} \textsc{cp}-stay  \\
    `Having stayed in all those houses' 
\ex
\gll {\em last}=ka  \textbf{baapa}  Badulla=nang   anà-pii \\
     last=\textsc{loc} father Badulla=\textsc{dat} \textsc{past}-go  \\
    `at last, father returned to Badulla.' (K051205nar05)
\z
\z
}

Next to kin terms, proper names can also be used as terms of address \xref{ex:pragm:address:propername1}\xref{ex:pragm:address:propername2}.



\xbox{16}{
\ea\label{ex:pragm:address:propername1}
\gll Andare=nang asà-panggel anà-biilang: `Se=dang \textbf{Andare}=pe biini=yang mà-caa\u nda suuka.' \\
      Andare=\textsc{dat} \textsc{cp}-call \textsc{past}-say \textsc{1s=dat} Andare=\textsc{poss} wife=\textsc{acc} \textsc{inf}-meet want\\
    `He called Andare and said: ``I want to meet Andare's wife''.' (K070000wrt05)
\z
}


\xbox{16}{
\ea\label{ex:pragm:address:propername2}
\gll \textbf{Aashik}=nang hathu {\em soldier} mà-jaadi suuka=si katha arà-caanya. \\
     Aashik=\textsc{dat} \textsc{indef} soldier \textsc{inf}-become like=\textsc{interr} \textsc{quot} \textsc{non.past}-ask  \\
    `He asks if you want to become a soldier, Ashik.' (B060115prs10)
\z
}
 
 
% \xbox{16}{
% \ea\label{ex:form:unreferenced}
% \gll Sebastian pùddas arà-maakang=si. \\
%       Sebastian spicy \textsc{non.past}-eat=\textsc{interr}
%     `Do you eat spicy food, Sebastian?' (B060115cvs02)
% \z
% }
 

Elder relatives may also be addressed  as or referred  to with the English terms \em uncle \em and \em auntie \em \xref{ex:pragm:address:uncle1}-\xref{ex:pragm:address:uncle3}. This does not imply that the person is actually a sibling of a parent.
 

\xbox{16}{
\ea\label{ex:pragm:address:uncle1}
\gll \textbf{\em Uncle}, kitham=pe  umma   anà-biilang  giini. \\
     uncle \textsc{1pl}=\textsc{poss} mother \textsc{past}-say like.this  \\
    `Uncle, our mother said the following: ... .' (K051220nar01)
\z
}
 

\xbox{16}{
\ea\label{ex:pragm:address:uncle2}
\gll Diya-la,  \textbf{{\em uncle}} se=ppe  SSC {\em exam} mà-ambel=nang duva-pul aari su-aada. \\
     see-\textsc{imp} uncle \textsc{1s=poss} SSC exam \textsc{inf}-take=\textsc{dat} two-ty day \textsc{past}-exist \\
    `See, uncle, there are (were) only twenty days left until my SSC exam.' (K051220nar01)
\z
}
 
\xbox{16}{
\ea\label{ex:pragm:address:uncle3}
\ea
\gll Thuan Kuddhuus, Thuan Thungku, Thuan Skiilan, Thuan Idriis,  \\
      Thuan Kuddhuus Thuan Thungku Thuan Skiilan Thuan Idriis \\
    `Thuan Kuddhuus, Thuan Thungku, Thuan Skiilan, Thuan Idriis,'  
\ex
\gll Thuan -- \textbf{{\em uncle}}, hathyang naama pada saapa ? \\
     Thuan { } uncle other name \textsc{pl} who  \\
    `Thuan ... --- what's the names of the others again, uncle?' (K060108nar02)
\z
\z
}

\subsection{Distancing}\label{sec:pragm:distancing}
Distancing oneself from the provided content can be done by \trs{sedang kalu}{as for me}, \trs{sedang thaau mosthornang}{as far as I know} or \trs{boolebiilang}{you could say that ...}


\xbox{16}{
\ea\label{ex:pragm:distancing:sedangkalu}
\gll Kaake=nang  dhraapa=so   bannyak aanak pada,  \textbf{se=dang} \textbf{kalu} blaangang thàrà-thaau. \\
     grandfather=\textsc{dat} how.many=\textsc{undet} much child \textsc{pl} \textsc{1s=dat} if amount \textsc{neg}-know  \\
    `Grandfather had several children, many children, as for me, I do not know the number.' (K051205nar05)
\z
}

\xbox{16}{
\ea\label{ex:pragm:distancing:sedangthaaumosthor}
\gll \textbf{Se=dang} \textbf{thaau} \textbf{mosthor=nang} inni=jo        inni     kithang=pe inni     Seelong=nang {\em political} {\em news}. \\
      \textsc{1s=dat} know manner=\textsc{dat} \textsc{prox}=\textsc{emph} \textsc{prox} \textsc{1pl}=\textsc{poss} \textsc{prox} Ceylon=\textsc{dat} political news \\
    `As far as I know, this is the political news for Sri Lanka.' (N061031nar01)
\z
}

\xbox{16}{
\ea\label{ex:pragm:distancing:bolebiilang}
\gll \textbf{Bole-biilang}    se=ppe    kaake=le            asàdhaathang hathu  {\em army} {\em officer} Badulla=ka. \\
      can-say \textsc{1s=poss} grandfather=\textsc{addit} \textsc{copula} \textsc{indef} army officer Badulla=\textsc{loc}\\
    `You can say that my grandfather also was an army officer in Badulla.' (B060115cvs05)
\z
}

Direct attribution of the source is also possible, as in \xref{ex:pragm:distancing:spaaruoorang}, where the source of the information \trs{spaaru oorang pada}{some people} is explicitly mentioned.

\xbox{16}{
\ea\label{ex:pragm:distancing:spaaruoorang}
\gll \textbf{Spaaru} \textbf{oorang} \textbf{pada} \textbf{arà-biilang} Seelong=nang  {\em English} anà-aaji.baa \textbf{katha} \\
     some man \textsc{pl} \textsc{non.past}-say Seelon=\textsc{dat} English \textsc{past}-bring.\textsc{anim} \textsc{quot}  \\
    `Some people say it is the  English who brought them to Sri Lanka.'  (K060108nar02)
\z
}






\subsection{Reinforcing}\label{sec:pragm:Reinforcing}
\citet{Slomanson2008ismil} reports the use of the string \em apa kata k$@$m-bilang \em (his orthography) for reinforcing purposes. The following is an example he gives.



\xbox{16}{
\ea
\gll Apa  kata   k$@$m-bilang, April ka  jo      e-datang (aDa) \\
     what word when-say    April  in FOC ASP-come  AUX  \\
    `What I say to you is that it has come in April.' \citep[7(17a)]{Slomanson2008ismil}
\z
}

With commands, \em -la \em \formref{sec:morph:-la} can be used as a polite reinforcer, while \em -de \em is a very impolite way to reinforce a command \formref{sec:morph:-de}.

% \section{Speech genres}\label{sec:pragm:Speechgenres}
% \section{Conversation}\label{sec:pragm:Conversation}
% How is conversation organized?
% 
% \subsection{Turn taking}\label{sec:pragm:Turntaking}
% How is turn taking organized?
% bukang indicates that another turn is to follow



% \xbox{16}{
% \ea
% \gll Kithang cumma aràduuduk \\
%        \\
%     `We are just waiting (bored).' (nosource)
% \z
% }

